---
title: "Public Health and Economic Impact of Hydro-Meteorological Events (Storms)"
author: "John James"
date: "March 6, 2016"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 12
    keep_md: yes
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
read_chunk("master.R")
read_chunk("load.R")
read_chunk("review.R")
read_chunk("process.R")
read_chunk("analyze.R")
read_chunk("present.R")
knitr::opts_knit$set(root.dir = 'C:/Users/John/Documents/GitHub/RepResearch_Course_Project_2')
knitr::opts_chunk$set(fig.width=12, fig.height=16, fig.path = "figures/", echo=TRUE, warning=FALSE)
```

## Synopsis

## Introduction
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

The purpose of this project was to explore and analyze the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to determine which types of hydro-meteorological events posed the greatest public health and economic cost.

## Research Question
The report seeks to address the following questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Data 
This report analyzed the NOAA Storm Events Database which documents the occurrence of storms and other significant weather phenomena having sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce. The database currently contains data from January 1950 to November 2015, as entered by NOAA's National Weather Service (NWS), capturing fatalities, property damage, and crop damage for some 48 designated event types.

## Data Processing
### Environment Setup
The following code configures the environment for the analysis.  Filenames are designated, the directory structure is established, and the requisite files and libraries are sourced.  

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<environment>>
```

### Load Data
This function downloads a zipped file from the web and stores it in the requested library. The function checks of the raw data directory exists, and creates it if it doesn't.  Next, the file is downloaded if the file doesn't exist, or the the file wasn't downloaded on the current day. Finally, we read the file into a data frame and return it to the master script.

  * Args:

    + url - the url for website from which the file will be downloaded
    + dir - the redirectory to which the file will be stored
    + zip - the name of the zip file
    + CSV - the name of the csv file
    + df  - the name of thd data frame that contains the data

  * Response: This function will load the file into a data frame and return it to the calling script
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<loadDataFunc>>
```

The function above is called here if "Load" (from environment variables) is set to TRUE
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<loadDataCall>>
```

### Data Review
The following code provides the dimensions of the dataset, the number of unique event types, and the number of rows with zero values for each of the variables being analyzed.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<reviewDataFunc>>
```

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<reviewData>>
```

There are **`r observations`** observations in the original data file, each having **`r variables`** variables.  The data set contains **`r numEventTypes`** unique event types and **`r zeroRows`** rows have zero values for each of the variables being measured.


### Data Transformations
Before analyzing the data, several data transformations were undertaken. 

#### Remove Zero Rows
This statement removes rows in which the value of each of the measurement variables is zero.

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<removeZeroRows>>
```

#### Extract Variables
The variables of interest for this analysis were:

* EVTYPE - The type of event.  The data contained 985 factor levels.
* BGN_DATE - The date upon which the weather event occurred
* FATALITIES - The number of fatalities directly caused by the weather event
* INJURIES - The number of persons injured as a direct consequence of the weather event.
* PROPDMG - The estimated amount of property damage indicated in dollar amounts adjusted by the PROPDMGEXP factor
* PROPDMGEXP - This factor indicates multiplicative factor to apply to the PROPDMG number.  For instance, a value of "K" indicates that the number in PROPDMG should be multiplied by 1000.  
* CROPDMG - The estimated amount of crop damage indicated in dollar amounts adjusted by the PROPDMGEXP factor
* CROPDMGEXP - This factor indicates multiplicative factor to apply to the CROPDMG number.  For instance, a value of "2" indicates that the number in PROPDMG should be multiplied by 100.  

The following code extracts the variables of interest for this analysis:

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<extractVariables>>
```

#### Event Type
There were 48 designated event types; however, there were 985 different factor levels for EVTYPE.  The following function and the associated call recodes the EVTYPE variable to one of the 48 designated values.  The designated event types are described in the NOAA Storm Data Directive NWSPD 10-16 which is available on the NOAA website at https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf. 

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<processEventTypeFunc>>
```

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<processEventTypeCall>>
```

#### Property & Crop Damage
The following function along with its function call, converts the property and crop damage estimates to dollar values. PROPDMGEXP and CROPDMGEXP contain the base ten exponent by which the property and crop damage estimates must be multiplied to reflect dollar terms.  Lastly, two new variables are created: one to hold the sum of fatalities & injuries and the second contains total damage which is the sum of property and crop damage estimates.

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<processDamageFunc>>
```

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<processDamageCall>>
```

#### Year
The following function, along with its function call, creates a new categorical variable that is equal to the year in which the event occurred.  This factor will be used to analyze health and economic impacts over time.


```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<processYearFunc>>
```

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<processYearCall>>
```

## Data Analysis
Health and economic impacts were evaluated from two perspective points of view. First, the health and economic effects were examined over the entire data set, by event type.  Next, time series plots were rendered to illuminate any trends over time. Finally, we observe impacts over a subset of the most recent data based upon any trends observed.

### Top 5 Event Types
The following six functions calculate the top 5 event types, in terms of health and economic impact.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<analysisFunc>>
```

The above functions are called to calculated top 5 event types in terms of health and economic impact, across all data.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<top5AllData>>
```

Next, we select storm data published since 1980.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<extract1980Data>>
```

Then we use the "top 5" functions to calculate top 5 event types, in terms of health and economic impact, for data published since 1980.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<top51980>>
```

Lastly, we calculate the time series data that will be used to reveal trends over time.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<timeSeriesData>>
```

### Presentation Functions
The following section contains the functions and calls required to present the data graphically.  

#### Bar Plot Functions
The following functions are used to render bar plots of the top 5 event types, in terms of health and economic impact.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<barPlotFuncs>>
```

#### Time Series Plot Functions
The following functions are used to render line charts of the top 5 event types, in terms of health and economic impact.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<trendLineFuncs>>
```

## Results
The following code renders a group of bar plots, showing the health and economic impacts for storm data from 1950.

```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<top5BarPlotsAll>>
```
Using the full data set as a source, it would appear that Tornados have the greatest public health and economic impact.

However, the following line charts reveal several interesting points.
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<trendLine>>
```
The charts reveal a lack of data for the period leading up to the 1980's.  After which, one can observe a wider range of weather events.  

The following chart illucidates the health and economic impacts since 1980.  
```{r echo = TRUE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=FALSE)}
<<top5BarPlots1980>>
```



Using this data as source, 


## References
(NOAA), U. N. (2016, March 8). NOAA Storm Database. Retrieved from NOAA National Centers for Environmental Information: https://www.ncdc.noaa.gov/stormevents/

